---
title: 'Chapter 04: Data Visualization'
author: "Robin Donatello"
date: "_Last update `r Sys.Date()`_"
output:
  html_document:
    toc: yes
    toc_float: true
  
---

# Load Libraries

```{r, echo=-2, warning=FALSE, message=FALSE}
library(scales); library(dplyr); library(gridExtra); 
library(gridBase); library(grid); library(ggplot2)
library(xtable); library(magrittr); library(knitr)
library(vcd)
opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, cache=TRUE)
```

# Helper functions
```{r}
univ.n.p <- function(df, var, dig){
  df %>% select_(var) %>% na.omit() %>% group_by_(var) %>% 
    summarise(n=n()) %>% 
    mutate(pct=round(n/sum(n),dig))
}

univ.n.pct <- function(df, var){
  df %>% select_(var) %>% na.omit() %>% group_by_(var) %>% 
    summarise(n=n()) %>% 
    mutate(pct=paste0(round(n/sum(n)*100,1), '%'))
}

biv.n.col.p <- function(df, row.var, col.var) {
  df %>% group_by_(row.var, col.var) %>% 
    summarise(count=n()) %>%
    mutate(pct=round(count/sum(count),3))
}
```

############################################
  
# Load & prepare data
```{r}
data.path <- "data6e/"

raw <- read.delim(paste0(data.path,"Parhiv.txt"), sep="\t", header=TRUE)

parhiv <- raw %>% 
  mutate(JOBMO = factor(JOBMO, 
                        labels=c("Employed", "Unemployed", "Retired/Disabled")), 
         EDUMO = factor(EDUMO, 
                        labels=c("Less than High School", "High School Graduate/GED", 
                                 "Post Secondary")),
         bsi_overall = rowMeans(raw[,grep("BSI", names(raw))]), 
         log_bsi_overall = log(bsi_overall+.01), 
         parent_bonding = (PB01+(4-PB02)+(4-PB04)+PB05+PB06+
                           PB11+PB12+(4-PB14)+(4-PB16)+PB17+(4-PB18)+(4-PB24))/12)
parhiv$AGEALC[parhiv$AGEALC==0] <- NA
parhiv$AGEMAR[parhiv$AGEMAR==0] <- NA
parhiv$AGESMOKE[parhiv$AGESMOKE==0] <- NA

raw.d <- read.delim(paste0(data.path,"Depress.txt"), sep="\t", header=TRUE)
depress <- raw.d %>% 
  mutate(gender = factor(SEX, 
                         labels=c("Male", "Female")), 
         educat =  factor(EDUCAT, 
                         labels=c("<HS", "Some HS", "HS Grad", "Some college", 
                                  "BS", "MS", "PhD")), 
         marital = factor(MARITAL, 
                         labels = c("Never Married", "Married", "Divorced", 
                                    "Separated", "Widowed")))

raw.l <- read.delim(paste0(data.path,"Lung.txt"), sep="\t", header=TRUE)

mice <- read.delim(paste0(data.path,"Mice.txt"), sep="\t", header=TRUE)
```

#### Demo 
```{r}
hist(parhiv$bsi_overall);box()

ggplot(parhiv, aes(x=bsi_overall)) + 
  geom_histogram() +
  theme_bw(base_size=16)

ggplot(parhiv, aes(x=bsi_overall)) + 
  geom_histogram() + 
  theme_classic(base_size = 20) + 
  xlab("BSI score") + ylab("Frequency") + ggtitle("Brief Symptom Inventory Score")
```

# Univariate

## Categorical 

#### Univariate barplot with counts
```{r}
ggplot(filter(parhiv, !is.na(EDUMO)), aes(x=EDUMO)) + 
  theme_classic(base_size = 20) + 
  geom_bar(aes(y = ..count..)) + ggtitle("Frequency of educational level") + 
  xlab("Highest educational level attained") +
  scale_x_discrete(labels = function(x) lapply(strwrap(x, width = 12, simplify = FALSE), 
                                               paste, collapse="\n")) + 
  geom_text(aes(y=..count.. + 5, label=..count..), stat='count', size = 7) + 
  theme(axis.ticks.x=element_blank()) 
```

#### Univariate barplot with percentages
```{r}
ptab.ctp <- univ.n.p(parhiv, "EDUMO", dig=3)

ggplot(ptab.ctp, aes(x=EDUMO)) + 
  geom_bar(aes(y=pct), stat="identity") + 
  xlab("Highest educational level attained") + ylab("") + 
  ggtitle("Percent of mothers with each \n educational level") + 
  theme_classic(base_size = 20) + 
  scale_y_continuous(labels=percent, limits=c(0,1)) + ylab("") + 
  scale_x_discrete(labels = function(x) lapply(strwrap(x, width = 12, simplify = FALSE), 
                                               paste, collapse="\n")) + 
  geom_text(aes(y=pct + .05, label=paste0(pct*100, "%")), size = 7) + 
  theme(axis.ticks.x=element_blank()) 
```


#### Cleveland Dotplot
```{r}
ptab.empl <- univ.n.pct(depress, "marital")

ggplot(ptab.empl, aes(x=n, y=reorder(marital, n), label=n)) +  
  geom_point(size = 3) + 
  geom_text(size=5, vjust=-1) + 
  xlab("Frequency") + ylab("Marital Status") + ggtitle("Frequency of marital status") +
  theme_classic(base_size=20) + 
  scale_x_continuous(limits=c(0,130))+
  theme(panel.grid.major.x = element_blank(), axis.ticks.y=element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color='grey60', linetype='dashed'))
```


#### Pie Chart
```{r}
dc <- table(depress$marital)

pie(dc, labels = paste0(names(dc), ' (', round(prop.table(dc)*100,1), "%)"), 
    col = c("#f7f7f7","#cccccc", "#969696", "#636363", "#252525"), cex=1.5, radius=0.5)
```

#### Stem-leaf plot
```{r}
par(mar=c(0,0,0,0), oma=c(0,0,0,0))
plot(c(0,0.5), c(0,1), type="n", axes=FALSE, main="", ylab="", xlab="")
tmp <- capture.output(stem(sort(depress$AGE)))
tmp <- tmp[4:length(tmp)]
text(0,1, paste(tmp, collapse='\n'), adj=c(0,1), family='mono', cex=1)
```

#### Stripchart
```{r}
ggplot(depress, aes(y=AGE, x=1)) + 
  geom_jitter(position = position_jitter(0.2)) + 
  ylab("Age") + 
  scale_x_continuous(limits=c(0,2)) +  
  theme_classic(base_size=20)+ xlab("")+
  theme(panel.grid.major.y = element_blank(), axis.ticks.y=element_blank(),
        panel.grid.minor.y= element_blank(), axis.text.y=element_blank()) + 
  coord_flip()
```


## Continuous 

#### Univariate histogram
```{r}
ggplot(depress, aes(x=AGE)) + 
  geom_histogram(colour="black", fill="grey80", binwidth=10) +  
  xlab("Age") + ylab("Frequency") + 
  ggtitle("Distribution of age") + 
  theme_classic(base_size = 16) +
  scale_x_continuous(limits=c(20,80), breaks=seq(20,80,by=10))
```

#### Univariate histogram with default bins 
```{r}
ggplot(depress, aes(x=AGE)) + 
  geom_histogram(colour="black", fill="grey80") + 
  xlab("Age") + ylab("Frequency") + 
  ggtitle("Distribution of age") + 
  theme_classic(base_size = 16) +
  scale_x_continuous(limits=c(15,90), breaks=seq(20,90,by=10))
```

#### Univariate Density
```{r}
ggplot(depress, aes(x=AGE)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="grey80") +  
  xlab("Age") + ylab("Density") + 
  geom_density(colour="black", lwd=1.1) + 
  scale_x_continuous(limits=c(15,90), breaks=seq(20,90,by=10)) + 
  ggtitle("Distribution of age") + theme_classic(base_size = 16)
```

#### Univariate boxplot
```{r}
txt <- boxplot(depress$AGE, plot=FALSE)$stats
data.labels <- data.frame(value=as.vector(txt), lab=as.character(txt))
labels <- data.frame(value=as.vector(txt), 
                     lab=c("Min", "Q(0.25)", "Median", "Q(0.75)", "Max"))

ggplot(depress, aes(y=AGE, x=1)) + 
  geom_boxplot(width=.3) + ylab("Age") + xlab("") + 
  scale_x_continuous(limits=c(0.5,1.35)) + 
  theme_classic(base_size = 16) +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) + 
  geom_text(data=data.labels,aes(x=0.75,y=value,label=lab),angle=0, size=5) + 
  geom_text(data=labels,aes(x=0.6,y=value,label=lab),angle=0, size=5)
```

#### Univariate boxplot modified
```{r}
txt <- boxplot(depress$CESD, plot=FALSE)$stats
data.labels <- data.frame(value=as.vector(txt), lab=as.character(txt))
labels <- data.frame(value=as.vector(txt), 
                     lab=c("Min", "Q(0.25)", "Median", "Q(0.75)", "Largest value \n within fence"))

ggplot(depress, aes(y=CESD, x=1)) + 
  geom_boxplot(width=.3) + ylab("CESD") + xlab("") + 
  scale_x_continuous(limits=c(0.75,1.5)) + 
  theme_classic(base_size = 16) + 
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) + 
  geom_text(data=data.labels,aes(x=1.25,y=value,label=lab),angle=0, size=5) + 
  geom_text(data=labels,aes(x=1.4,y=value,label=lab),angle=0, size=5)
```


#### Univariate boxplot
```{r}
ggplot(depress, aes(y=AGE, x=1)) + 
  geom_boxplot(width=.3) + geom_violin(alpha=.2) +
  ylab("Age") + xlab("") + theme_classic(base_size = 16) + 
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) + 
  stat_summary(fun.y=mean, geom="point", shape=18, size=4) 

ggplot(depress, aes(y=CESD, x=1)) + 
  geom_boxplot(width=.3) + geom_violin(alpha=.2) +  
  ylab("CESD") + xlab("") + theme_classic(base_size = 16) + 
  geom_jitter(alpha=.3, width=.1) +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) 
```

# Bivariate 

## Categorical ~ Categorical 

#### Two-way Frequency table
```{r}
table(depress$gender, depress$educat) %>% kable()
```

#### Two-way cell proportions
```{r}
kable(
  addmargins(
    prop.table(table(depress$gender, depress$educat))*100
    ),
  digits=1)
```

#### Two-way row proportions
```{r}
kable(
  addmargins(
    prop.table(table(depress$gender, depress$educat), margin=1)*100
  ),
  digits=1)
```

#### Two-way column proportions
```{r}
kable(
  addmargins(
    prop.table(table(depress$gender, depress$educat), margin=2)*100
  ),
  digits=1)
```

#### Bivariate stacked barchart
```{r}
phna <- parhiv %>% select(JOBMO, EDUMO) %>% na.omit()
pct.cc <- biv.n.col.p(phna, "EDUMO", "JOBMO")

ggplot(phna, aes(x=EDUMO, fill=JOBMO)) + 
  geom_bar() + 
  theme_classic(base_size = 16) + 
  scale_fill_grey(name="Job Status") + xlab("") + 
  theme(legend.position="top", legend.title=element_text(size=10), legend.text=element_text(size=10)) +
  coord_equal(1/30) + ylab("Frequency") + 
  scale_x_discrete(labels = function(x) lapply(strwrap(x, width = 12, simplify = FALSE), 
                                               paste, collapse="\n")) + 
  theme(axis.ticks.x=element_blank()) 
```

#### Bivariate stacked barchart
```{r}
ggplot(pct.cc, aes(x=EDUMO, fill=JOBMO)) + 
  geom_bar(aes(y=pct), stat='identity') + 
  scale_fill_grey(guide=FALSE) + 
  theme_classic(base_size = 16) + 
  xlab("") + ylab("Percent") + 
  scale_y_continuous(labels=percent) + 
  coord_equal(1/0.35) + 
  scale_x_discrete(labels = function(x) lapply(strwrap(x, width = 12, simplify = FALSE), 
                                               paste, collapse="\n")) + 
  theme(axis.ticks.x=element_blank()) 
```

#### Bivariate side by side barchart
```{r}
ggplot(pct.cc, aes(x=EDUMO, y=pct, fill=JOBMO)) + 
  scale_fill_grey(name="Job Status") + 
  geom_bar(stat="identity", position = "dodge") +  
  theme_classic(base_size = 16) + xlab("") + 
  scale_y_continuous(labels = percent) + 
  ylab("Percent") + 
  scale_x_discrete(labels = function(x) lapply(strwrap(x, width = 12, simplify = FALSE), 
                                               paste, collapse="\n")) + 
  theme(axis.ticks.x=element_blank()) 
```

#### Bivariate dot plot for differences
```{r}
ggplot(pct.cc, aes(x=count, y=reorder(JOBMO, count))) + 
  geom_point(size = 3) + 
  xlab("Frequency") + 
  facet_grid(EDUMO ~ ., labeller=label_wrap_gen(14)) +
  geom_segment(aes(yend=JOBMO), xend=0, color='grey50') + 
  ylab("Job Status") + theme_bw(base_size = 16) +
  theme(panel.grid.major.y = element_blank())  + 
  theme(axis.ticks.y=element_blank()) 
```


#### Bivariate dot plot for differences
```{r}
bdpd <- parhiv %>% mutate(gender=factor(GENDER, labels=c("Male", "Female"))) %>% 
           select(JOBMO, gender) %>% na.omit() %>% 
           group_by(JOBMO, gender) %>% 
           summarise(count=n())

ggplot(bdpd, aes(x=count, y=reorder(JOBMO, count))) + 
              geom_point(aes(shape=gender), size=3) + 
              geom_line(aes(group=JOBMO)) +
  scale_shape_discrete(name="Gender of \n teen") +
  xlab("Frequency") + 
  ylab("Mother's Job Status") + 
  theme_bw(base_size = 16) +
  theme(panel.grid.major.y = element_blank())  + 
  theme(axis.ticks.y=element_blank()) 
```


####  Mosaic Plot
```{r}
kable(
  prop.table(
    table(phna$EDUMO, phna$JOBMO)
    )*100, 
  digits=1)

varnames <- list(set_varnames=c(JOBMO="Job Status", EDUMO="Highest Education Level"), 
                 rot_labels = c(0,90,0,0),
                 offset_varnames = c(left = 5, top=1), 
                 just_labels=c(left="right"))
lnames <- list(EDUMO = c("Less than\nHigh School", "High School\nGraduate/GED", "Post\nSecondary"))

mosaic(JOBMO~EDUMO, data=phna, color=TRUE, 
    set_labels=lnames,
    labeling_args = varnames,  
    margins=c(top=5, right=0, left=7))
```

## Continuous ~ Continuous 

#### Scatterplot 
```{r}
ggplot(depress, aes(x=AGE, y=CESD)) + geom_point(size=1, col="grey") + theme_classic(base_size = 16) +  
  xlab("Age") + ylab("CESD") + 
  geom_smooth(se=FALSE, method="lm", color="black", aes(linetype="Linear"))+ 
  geom_smooth(se=FALSE, method="loess", color="black", aes(linetype="Lowess"))+ 
  scale_linetype_manual(values = c("solid","dashed"), name="line types",
                        labels=c("Linear","Lowess"),guide="legend")
```


#### Profile plot 
```{r}
ggplot(mice, aes(y=WEIGHT, x=DAY, group=ID)) + 
  geom_point() + geom_line() + 
  theme_classic(base_size = 16) + 
  ylab("Weight (g)") + xlab("Time (Days)")
```

## Bivariate Continuous + Categorical 
```{r}
ggplot(depress, aes(y=CESD, x=gender, color=gender)) +
  scale_color_grey(start=0, end=0.5,guide=FALSE) +
  geom_boxplot(width=.3) + geom_violin(alpha=.2) +  
  ylab("CESD") + xlab("") + 
  theme_classic(base_size = 20) + 
  stat_summary(fun.y=mean, geom="point", shape=18, size=4)  + 
  theme(axis.ticks.x=element_blank()) 

ggplot(depress, aes(x=CESD, color=gender)) + 
  geom_density() + theme_classic(base_size = 20) + 
  ylab("Density") + xlab("CESD")+
  scale_color_grey(start=0, end=0.5)+
  theme(axis.ticks = element_blank(), axis.text.y = element_blank()) + 
  theme(legend.justification=c(1,1), legend.position=c(1,1))

ggplot(depress, aes(x=CESD, fill=gender)) +  
  facet_grid(gender ~ .) + 
  xlab("CESD")+ylab("Frequency") + 
  geom_histogram() + theme_bw(base_size = 20) + 
  scale_fill_grey(start=0, end=0.5,guide=FALSE) +
  theme(axis.ticks = element_blank(), axis.text.y = element_blank()) + 
  theme(legend.justification=c(1,1), legend.position=c(1,1))
```


# Multivariate

#### Scatterplots 
```{r}
ggplot(depress, aes(x=AGE, y=CESD, color=gender)) + 
  scale_color_grey() + 
  geom_point() +  
  theme_classic(base_size = 20) + 
  xlab("Age") + ylab("CESD")

ggplot(depress, aes(x=AGE, y=CESD, shape=gender)) + 
  geom_point() +  
  theme_classic(base_size = 20) + 
  xlab("Age") + ylab("CESD")

ggplot(depress, aes(x=AGE, y=CESD, size=INCOME)) + 
  geom_point(shape=21) + 
  theme_classic(base_size = 20) + 
  xlab("Age") + ylab("CESD") + 
  scale_size_continuous(limits=c(0,70), breaks = seq(0,70,by=10))

ggplot(depress, aes(x=AGE, y=CESD, fill=INCOME)) + 
  geom_point(shape=21, size=5) +
  theme_classic(base_size = 20) + 
  xlab("Age") + ylab("CESD") + 
  scale_fill_gradient(low="white", high="black")
```


#### Hist panel
```{r}
ggplot(filter(parhiv, !is.na(EDUMO) & !is.na(JOBMO)), aes(x=bsi_overall)) +  
  ylab("Frequency") + 
  geom_histogram()+  
  facet_grid(EDUMO~JOBMO, labeller=label_wrap_gen(14)) +
  scale_y_continuous(breaks=seq(0,10,by=2))+
  theme_bw(base_size = 18) +
  xlab("Overall BSI score")

```

#### Scatterplot matrix
```{r}
library(psych)
c.vars <- parhiv[,c("bsi_overall", "parent_bonding", 'AGESMOKE', 'AGEALC')]
names(c.vars) <- c("BSI", "Parental care subscale", "Age started smoking", "Age started drinking")

pairs.panels(c.vars, hist.col="grey80", col=1, rug=FALSE, cex.cor = .5, ellipses = FALSE)
```


#### Correlation plot
```{r}
library(corrplot)

corrplot(cor(c.vars, use='pairwise.complete.obs'), type="upper", col=gray.colors(100), 
         tl.col="black", tl.srt=30)
```

